---
title: "MHickey Covid-19 Dashboard"
output: html_document
runtime: shiny
---

<style type="text/css">
.main-container {
  max-width: 1250px;
  margin-left: auto;
  margin-right: auto;
}
</style>


## Visualizing the Spread and Impact of Covid-19 in the United States

This notebook represents a series of graphics that assess the impact of Covid-19
at a global, national, regional, and state level in terms of reported PCR testing, cases, hospitalizations,
and deaths. This is intended as a retrospective look at the data - *not* as a model to project future spread
or impact of the virus.

There are a number of data sources leveraged to create these graphics:

- The Covid Tracking Project 
  ("https://covidtracking.com/")
  
- US State population figures 
  ("https://www.census.gov/data/tables/time-series/demo/popest/2010s-state-total.html#par_textimage_1574439295")
  
- John's Hopkins Global Covid Death Data
  ("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/")
  
  
Note that data quality has not been perfect, especially in the early months of the pandemic. Reports are
aggregated from hundreds of municipalities, all of which have different practices. When interpreting the data, keep in mind the following:

- Reported Covid-19 cases represent only a fraction of all infections
- Reported deaths tend to lag the trends of other metrics by several weeks
- No single metric provides a perfect representation of the situation

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 360)
# print(getwd())
```

Load relevant packages:
```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Load libraries
# Use `install.packages("package_name") if package has not yet been installed

library(tidyverse) # For general data wrangling and cleaning
library(shiny)     # For interactive visuals
library(ggthemes)  # For preset graphic theme settings
library(gganimate)
library(zoo)
library(grid)
library(scales)
library(lubridate) # For converting time values

# For mapping data
library(sf)                   
library("rnaturalearth")
library("rnaturalearthdata")
library(rgeos)

```


```{r, echo = FALSE, message = FALSE, warning = FALSE}

# Connect to APIs, and wrangle data

# Read in data from The Covid Tracking Project
# This dataset contains detailed daily data on PCR testing, 
# hospitalizations, and deaths for each of the 50 US States plus the District
# of Columbia.

CovidData <- read_csv("https://covidtracking.com/api/v1/states/daily.csv") %>%
  mutate(date = as.Date(as.character(date),format="%Y%m%d"),
         NewPercentPositive = 100*positiveIncrease/(totalTestResultsIncrease)) %>%
  filter(NewPercentPositive >= 0) %>%
  transmute(date, state, totalTestResultsIncrease, positiveIncrease, negativeIncrease, 
            hospitalizedCurrently, deathIncrease, death)

bak <- CovidData

Pop <- read_csv("Statepop.csv")

CovidData <- left_join(CovidData,
                       Pop,
                       by = "state")

# Map to region
Region <- data.frame(state = c("AK", "HI", "WA", "OR", "CA", "NV", "ID", "MT", "WY", "UT", "AZ", "CO", "NM",
                               "ND", "SD", "NE", "KS", "MN", "IA", "MO", "WI", "IL", "IN", "MI", "OH",
                               "TX", "OK", "AR", "LA", "KY", "TN", "AL", "MS", "WV", "VA", "DC", "MD",
                               "DE", "NC", "SC", "GA", "FL",
                               "PA", "NJ", "NY", "CT", "MA", "VT", "NH", "ME", "RI"),
                     region = c("West", "West", "West", "West", "West", "West", "West", "West", "West", "West", "West",
                                "West", "West",
                               "Midwest", "Midwest", "Midwest", "Midwest", "Midwest", "Midwest", "Midwest", "Midwest", 
                               "Midwest", "Midwest", "Midwest", "Midwest",
                               "South", "South", "South", "South", "South", "South", "South", "South", 
                               "South", "South", "South", "South",
                               "South", "South", "South", "South", "South",
                               "Northeast", "Northeast", "Northeast", 
                               "Northeast", "Northeast", "Northeast", "Northeast", "Northeast", "Northeast"))

CovidData <-
  left_join(CovidData,
            Region,
            by = "state") %>%
  filter(!(is.na(region)))

CovidData$region <- factor(CovidData$region, 
                           levels = c("Northeast", "Midwest", "South", "West"))

PlotBase <- ggplot(CovidData) +
  theme_tufte() +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  labs(caption = "Data Source: The Covid Tracking Project") +#\nPlot created by @KeepTheOutliers") +
  theme(axis.text = element_text(size = 14, family = "serif"),
        axis.title = element_text(size = 16, family = "serif"),
        legend.title = element_text(size = 14, family = "serif"),
        legend.position = "top",
        plot.title = element_text(size = 18, family = "serif", hjust = 0.4),
        plot.background = element_rect(fill = "#faf8f6"),
        axis.ticks.y = element_blank(),
        panel.ontop = TRUE,
        panel.grid.major.y = element_line(color = "#faf8f6", size = 0.5),
        panel.grid.minor.y =  element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.caption = element_text(colour = "grey25", family = "serif", hjust = 0.5, size = 10)) +
  scale_x_date(limits = c(as.Date("2020-03-01"), as.Date(Sys.Date()))+1,
               date_breaks = "1 month", date_labels = "%b")

# CovidData 
```

# Global Covid-19 Impact

The figure below maps the reported Covid-19 deaths per capita across the world. The countries
with the highest reported deaths per capita are in Europe and the Americas.

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=12, fig.height = 8}

# Pull in world population data by country
world <- ne_countries(scale = "medium", returnclass = "sf")

# Generate string of today's date
Today <- paste0(substr(Sys.Date()-1, 6, 10), "-", substr(Sys.Date()-1, 1, 4))

# Connect to API with John's Hopkins global Covid-19 death report data
# raw <- read_csv(file = paste0("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/",
#                               "csse_covid_19_data/csse_covid_19_daily_reports/",
#                               Today, ".csv"))
raw <- read_csv(file = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/12-06-2020.csv")

raw %>%
  mutate(Date = as_date(Last_Update)) %>%
  group_by(Province_State, Date) %>%
  summarize(DailyDeaths = sum(Deaths)) %>%
  ungroup() -> CovidDataProvince

raw %>%
  mutate(Date = as_date(Last_Update),
         Country_Region = ifelse(Country_Region == "US", "United States of America", Country_Region)) %>%
  group_by(Country_Region, Date) %>%
  summarize(DailyDeaths = sum(Deaths)) %>%
  ungroup() -> CovidDataCountry


MapData <-
  left_join(world,
            CovidDataProvince,
            by = c("admin" = "Province_State")) %>%
  filter(!(is.na(DailyDeaths))) %>%
  rbind(left_join(world,
                  CovidDataCountry,
            by = c("sovereignt" = "Country_Region"))) %>%
  mutate(DeathsperMillion = DailyDeaths/pop_est*100000)

LatestDate <- max(CovidData$date)

# Plot World
ggplot(data = MapData) +
    geom_sf(aes(fill = DeathsperMillion),
            size = 1/4) +
  coord_sf(crs = "+proj=moll") +
  theme_map() +
  labs(fill = "Reported Covid Deaths per Million Residents") +
  # theme(legend.position = "bottom") +
  labs(title = "Total Covid Deaths per Million",
       subtitle = paste0("as of ", LatestDate)) +
  theme(legend.text = element_text(family = "serif", size = 12, color = "grey25"),
        legend.title = element_text(family = "serif", size = 14, color = "grey25"),
        plot.title = element_text(family = "serif", size = 28, hjust = 0.5, color = "grey25"),
        plot.subtitle = element_text(family = "serif", size = 14, hjust = 0.5, color = "grey25")) +
  scale_fill_continuous(low = "#FFF4D9", high = "dark red",
                        limits = c(0, 200), na.value="grey95") -> map

map

```

```{r, echo = F, fig.width=12, fig.height=6}
MapData %>%
  as_data_frame() %>%
  filter(pop_est > 1000000) -> BoxPlotData

BoxPlotData$region_un <- factor(BoxPlotData$region_un,
                                levels = c("Americas", "Europe", "Asia", "Africa", "Oceania"))

arrow3 <-
  tibble(
    x1 = c("Africa"),
    x2 = c("Africa"),
    y1 = c(70),
    y2 = c(10)
  )


ggplot(BoxPlotData) +
  geom_boxplot(aes(x = region_un,
                   y = DeathsperMillion,
                   fill = region_un),
               width = 1/4,
               alpha = 1/3,
               outlier.alpha = 0)  +
  geom_jitter(aes(x = region_un,
                 y = DeathsperMillion,
                 color = region_un),
               size = 1.2,
              height = 0,
              width = 1/5,
              alpha = 2/4)  +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0, 125)) +
  theme_tufte() +
  labs(y = "Reported Covid-19 Deaths\nper Million",
       x = "World Region",
       title = "Europe and the Americas have\nReported the Highest Covid-19 Death Rates",
       fill = "Region",
       caption = "Data Source: The Covid Tracking Project") +#\nPlot created by @KeepTheOutliers") +
  annotate("text", x = "Africa", y = 90,
           family = "serif", size = 3.25, color = "grey25",
           label = "Africa's low death totals\nmay owe to its relatively\nyoung population") +
  geom_curve(data = arrow3,
             aes(x = x1, y = y1, xend = x2, yend = y2),
             size = 0.3, color = "gray45", curvature = 0.1) +
  theme(axis.text.y = element_text(size = 12, family = "serif"),
        axis.text.x = element_text(size = 12, family = "serif"),
        axis.title.x = element_text(size = 16, family = "serif"),
        axis.title.y = element_text(size = 16, family = "serif"),
        legend.title = element_text(size = 10, family = "serif"),
        legend.position = "top",
        plot.title = element_text(size = 18, family = "serif", hjust = 0.5),
        plot.subtitle = element_text(size = 12, family = "serif", hjust = 0.5),
        plot.background = element_rect(fill = "#faf8f6"),
        strip.text = element_text(size = 10, family = "serif", color = "grey25"),
        axis.ticks.y = element_blank(),
        panel.grid.minor.y =  element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.margin=unit(c(0.5,0.25,0.25,0.5),"cm"),
        panel.spacing.y = unit(1, "lines"),
        plot.caption = element_text(colour = "grey25", family = "serif", hjust = 0.5, size = 10)) +
  guides(fill = F, color = F)
```


# National Overview of Reported PCR Covid-19 Testing

The plot below shows all reported daily PCR tests in the United States since the
beginning of the Covid-19 pandemic. Positive tests are shown in orange, while 
negative tests are shown in green. The red line indicates the 7-day running average
of the "Positive Test Rate," or the percentage of all tests that return a positive
result.

Some high-level takeaways: 

- Testing capacity has increased continually since March

- Differences in testing capacity over time obscure the true
virus prevalence during different peaks

- Positive rate is at its highest value since early May

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=12, fig.height=6, dpi = 500}
options(scipen = 1000000)

arrowNatTest <- 
  tibble(
    x1 = c(as_date("2020-05-09")),
    x2 = c(as_date("2020-04-27")),
    y1 = c(1100000), 
    y2 = c(1000000),
    region = "Northeast"
  )

# Derive peak daily tests (rounded up to nearest 200,000 tests)
UsPeak <- (((((CovidData %>%
  filter(date == max(date)))$totalTestResultsIncrease %>% 
    sum())/200000) %>% 
  ceiling())*200000)

PlotBase +
  geom_bar(data = CovidData %>%
             pivot_longer(cols = c(positiveIncrease, negativeIncrease),
                          names_to = "Test Result",
                          values_to = "Increase") %>%
             mutate(`Test Result` = ifelse(`Test Result` == "positiveIncrease", "Positive", "Negative")),
           aes(x = date, y = Increase, fill = `Test Result`),
           stat = "identity", alpha = 1/2, width = 0.7) +
  labs(y = "Daily New Tests\n",
       x = "",
       fill = "Test Result",
       title = "Daily New Covid-19 Tests: United States") +
  geom_line(data = CovidData %>%
              filter(lubridate::as_date(date) > lubridate::as_date("2020-03-15")) %>%
              group_by(date) %>% 
              summarize(DailyPercentPositive = 100*sum(positiveIncrease)/sum(totalTestResultsIncrease)),
            aes(x = date,
                y = rollmean(DailyPercentPositive*UsPeak/25, 
                             7, na.pad=TRUE, align = "right")),
            color = "#a32d02",
            size = 1.25,
            alpha = 2/3) + 
  annotate("text", x = as_date("2020-05-20"), y = 1220000,
           family = "serif", size = 4.5, color = "#a32d02",
           label = "Positive Test Rate (%)\n[7-Day Average]") +
  geom_curve(data = arrowNatTest,
             aes(x = x1, y = y1, xend = x2, yend = y2),
             size = 0.4,
             color = "grey25", curvature = -0.2) +
  theme(axis.text.y.right = element_text(color = "#a32d02"),
        axis.title.y.right = element_text(color = "#a32d02")) +
  scale_y_continuous(label = comma,
                     sec.axis = sec_axis( trans=~.*20/2000000, name=""),#Positive Rate (%)\n(7-Day Running Average)\n"),
                     limits = c(0, UsPeak)) -> USTestPlot

USTestPlot
```


Below is a plot of the same data, but grouped into four regions:

- Northeast

- Midwest

- South

- West

We can see that the virus has spread has not been consistent across all regions, but that
positive tests and positive rate are climbing in every region as we enter the winter season.


```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=12, fig.height=6, dpi=300}
RegPeak <- ((((CovidData %>%
  filter(date == max(date)) %>%
  group_by(region) %>%
  summarize(Peak = sum(totalTestResultsIncrease)))$Peak %>% max())/50000) %>%
  ceiling())*50000
# Regional Breakdown
PlotBase +
  geom_bar(data = CovidData %>%
             pivot_longer(cols = c(positiveIncrease, negativeIncrease),
                          names_to = "Test Result",
                          values_to = "Increase") %>%
             mutate(`Test Result` = ifelse(`Test Result` == "positiveIncrease", "Positive", "Negative")),
           aes(x = date, y = Increase, fill = `Test Result`),
           stat = "identity", alpha = 1/2, width = 0.99995) +
  labs(y = "Daily New Tests\n",
       x = "",
       fill = "Test Result",
       title = "Daily New Covid-19 Tests: United States") +
  geom_path(data = CovidData %>%
              filter(lubridate::as_date(date) > lubridate::as_date("2020-04-01")) %>%
              ungroup() %>%
              group_by(date, region) %>% 
              summarize(DailyPercentPositive = 100*sum(positiveIncrease)/sum(totalTestResultsIncrease)) %>%
              arrange(region, date),
            aes(x = date,
                y = rollmean(DailyPercentPositive*RegPeak/30, 
                             7, na.pad=TRUE, align = "right"),
                group = region),
            color = "#a32d02",
            size = 1/2,
            alpha = 2/3) + 
  scale_y_continuous(label = comma,
                     sec.axis = sec_axis( trans=~.*30/600000, name="Positive Rate (%)\n(7-Day Running Average)\n"),
                     limits = c(0, RegPeak)) +
  theme(axis.text.y.right = element_text(color = "#a32d02", size = 12),
        axis.title.y.right = element_text(color = "#a32d02", size = 12),
        axis.text.x = element_text(angle = 90, size = 12),
        panel.grid.major.y = element_line(size = 1/6),
        panel.spacing = unit(1.5, "lines"),
        strip.text = element_text(family = "serif", size = 12, color = "grey25")) +
  facet_wrap(~region, nrow = 1, strip.position="top") +
  guides(color = F) -> RegionTestPlot

RegionTestPlot

# ggsave(filename = "RegionTestPlot.png",
#        plot = RegionTestPlot,
#        device = "png",
#        path = "/home/mhickey/Covid/")
```


# National Overview of Reported Hospitalizations due to Covid-19

Hospitalization Data
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=12, fig.height=6, dpi = 500}

arrow <- 
  tibble(
    x1 = c(as_date("2020-06-28")),
    x2 = c(as_date("2020-07-09")),
    y1 = c(50000), 
    y2 = c(48000),
    region = "South"
  )

PlotBase +
  geom_bar(aes(x = date, y = hospitalizedCurrently, fill = region),
           stat = "identity", alpha = 1/2, width = 0.6) +
  labs(x = "",
       y = "Current Hospitalizations\ndue to Covid-19\n",
       fill = "Region (per US Census)",
       title = "Covid-19 Hospitalizations by Region\n") +
  # geom_label(data = Annot,
  #          aes(x = x, y = y, label = text),
  #          family = "serif", size = 3) +
  annotate("text", x = as_date("2020-06-12"), y = 50000,
           family = "serif", size = 3.5, color = "gray25",
           label = "Florida begins reporting \nhospitalization data") +
  geom_curve(data = arrow,
             aes(x = x1, y = y1, xend = x2, yend = y2),
             arrow = arrow(length = unit(0.07, "inch")), size = 0.4,
             color = "gray45", curvature = -0.2) +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_line(size = 1/3)) -> HospStackPlot

# HospStackPlot


PlotBase +
  geom_bar(aes(x = date, y = hospitalizedCurrently, fill = ""),
           stat = "identity", alpha = 1/2, width = 0.6) +
  labs(x = "",
       y = "Current Hospitalizations\ndue to Covid-19\n",
       title = "US Covid-19 Hospitalizations are Higher than ever before\n") +
  annotate("text", x = as_date("2020-06-12"), y = 50000,
           family = "serif", size = 3.5, color = "gray25",
           label = "Florida begins reporting \nhospitalization data") +
  geom_curve(data = arrow,
             aes(x = x1, y = y1, xend = x2, yend = y2),
             arrow = arrow(length = unit(0.07, "inch")), size = 0.4,
             color = "gray45", curvature = -0.2) +
  # geom_label(data = Annot,
  #          aes(x = x, y = y, label = text),
  #          family = "serif", size = 3) +
  guides(fill = F)+
  theme(legend.position = "bottom",
        panel.grid.major.y = element_line(size = 1/3)) -> HospStackPlot2

HospStackPlot2

```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=12, fig.height=6, dpi = 300}
arrow3 <- 
  tibble(
    x1 = c(as_date("2020-06-21")),
    x2 = c(as_date("2020-07-08")),
    y1 = c(26000), 
    y2 = c(25000),
    region = "South"
  )

text <- 
  tibble(
    x = as_date("2020-05-01"),
    y = 26000,
    label = "Florida begins\n reporting\nhospitalization\ndata",
    region = "South"
  )
  

PlotBase +
  geom_bar(data = CovidData,
           aes(x = date, y = hospitalizedCurrently, fill = region),
           stat = "identity", alpha = 1, width = 1.1) +
  labs(x = "",
       y = "Current Hospitalizations\ndue to Covid-19\n",
       fill = "Region (per US Census)",
       title = "Hospitalizations Due to Covid-19 are Increasing across the US\n") +
  # geom_label(data = Annot,
  #          aes(x = x, y = y, label = text),
  #          family = "serif", size = 3) +
  # geom_text(data = text,
  #           aes(x = x, y = y, label = label),
  #           font = "serif", size = 3, color = "grey25") +
  # geom_curve(data = arrow3, 
  #            aes(x = x1, y = y1, xend = x2, yend = y2),
  #            arrow = arrow(length = unit(0.07, "inch")), size = 0.4,
  #            color = "gray45", curvature = -0.2) +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_line(size = 1/5)) +
  facet_wrap(~region, nrow = 1) +
  guides(fill = F) +
  scale_x_date(limits = c(as.Date("2020-03-01"), as.Date(Sys.Date()))+1,
               date_breaks = "2 months", date_labels = "%b") +
  theme(panel.spacing = unit(2.75, "lines"),
        plot.margin=unit(c(0.4,1,0.4,1.2),"cm"),
        axis.text.x = element_text(angle = 90, size = 12)) -> HospFacetPlot

HospFacetPlot


# ggsave(filename = "HospFacetPlot.png",
#        plot = HospFacetPlot,
#        device = "png",
#        path = "/home/mhickey/Covid/")

```


View Hospitalization Data by US State
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height= 6, dpi = 600}

options(scipen = 999)

HospPerCapita <- bak %>%
  filter(!(state %in% c("AS", "MP", "PR"))) %>%
  left_join(Pop,
            by = "state") %>%
  filter(!is.na(Pop)) %>%
  rowwise() %>%
  mutate(HospPerCapita = hospitalizedCurrently/Pop) %>%
  ungroup() %>%
  filter(!is.na(HospPerCapita)) %>%
  filter(HospPerCapita > 0) %>%
  left_join(Region,
            by = "state") %>%
  group_by(state)

Current <- HospPerCapita %>%
  group_by(state) %>%
  mutate(maxH = max(hospitalizedCurrently/Pop)) %>%
  filter(date == max(date)) %>%
  ungroup()


OrderH <- HospPerCapita %>%
  group_by(state) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  arrange(-HospPerCapita) %>%
  transmute(state) %>%
  unique()

HospPerCapita$state <- factor(HospPerCapita$state, 
                               levels = OrderH$state)
Current$state <- factor(Current$state, 
                        levels = OrderH$state)

ggplot() +
  # scale_color_brewer(palette = "Dark2") +
  # scale_color_manual(values = c("#a32d02")) +
  facet_wrap(~state, nrow = 3, strip.position="bottom") +
  # scale_y_log10() +
  scale_y_continuous(breaks = c(0, 200, 400, 600, 800, 1000, 1200)) +
  coord_cartesian(ylim = c(0, 1201)) +
  scale_x_date(limits = c(as.Date("2020-02-20"), Sys.Date()+3),
               breaks = seq(lubridate::as_date("2020-03-01"), lubridate::as_date(Sys.Date()+3), 62),
               date_labels = "%b") +
  theme_tufte() +
  labs(y = "Hospitalizations\nper Million\n",
       x = "",
       title = "Covid-19 Hospitalizations are Rising in nearly every US State",
       subtitle = "By US State",
       fill = "Region",
       caption = "Data Source: The Covid Tracking Project") +#\nPlot created by @KeepTheOutliers") +
  theme(axis.text.y = element_text(size = 12, family = "serif"),
        axis.text.x = element_text(size = 8, family = "serif", angle = 90),
        axis.title.x = element_text(size = 16, family = "serif"),
        axis.title.y = element_text(size = 16, family = "serif"),
        legend.title = element_text(size = 10, family = "serif"),
        legend.position = "top",
        plot.title = element_text(size = 18, family = "serif", hjust = 0.5),
        plot.subtitle = element_text(size = 12, family = "serif", hjust = 0.5),
        plot.background = element_rect(fill = "#faf8f6"),
        strip.text = element_text(size = 10, family = "serif", color = "grey25"),
        axis.ticks.y = element_blank(),
        panel.grid.minor.y =  element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.margin=unit(c(0.5,0.25,0.25,0.5),"cm"),
        panel.spacing.y = unit(1, "lines"),
        plot.caption = element_text(colour = "grey25", family = "serif", hjust = 0.5, size = 10)) -> HospPerCapitaPlot

topstate <- OrderH$state[1]

HospPerCapitaPlot +
  geom_bar(data = HospPerCapita,
              aes(x = date,
                  y = HospPerCapita*1000000,
                  fill = region),
           width = 1.5,
           alpha = 3/3,
           stat = "identity"
              # color = "#a32d02",
              ) +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_continuous(breaks = c(0:3)*400) +
  # scale_y_log10(limits = c(0.1, 100)) +
  scale_x_date(limits = c(as.Date("2020-03-01"), as.Date(Sys.Date())+1),
               date_breaks = "3 months", date_labels = "%b") -> HospPlot


unique(HospPerCapita$state) %>% length
```


```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi = 500, fig.width=12, fig.height=6}

# Define UI for app that draws a histogram ----
ui <- fluidPage(

  # App title ----
  # titlePanel("Hello Shiny!"),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar panel for inputs ----
     selectInput("state", "State:",
                 c("All US States", as.character(unique(HospPerCapita$state)) %>% sort()),
                 tableOutput("data")),
    # Main panel for displaying outputs ----
    mainPanel(

      # Output: Histogram ----
      plotOutput(outputId = "distPlot")

    )
  )
)
# Define server logic required to draw a histogram ----
server <- function(input, output) {

  # Histogram of the Old Faithful Geyser Data ----
  # with requested number of bins
  # This expression that generates a histogram is wrapped in a call
  # to renderPlot to indicate that:
  #
  # 1. It is "reactive" and therefore should be automatically
  #    re-executed when inputs (input$bins) change
  # 2. Its output type is a plot
  output$distPlot <- renderPlot({

    if(input$state == "All US States") {
      HospPlot
    } else{

    HospPerCapitaPlot +
    geom_bar(data = HospPerCapita %>% filter(state == input$state),
              aes(x = date,
                  y = HospPerCapita*1000000,
                  fill = region),
           alpha = 3,
           width = 2.5,
           stat = "identity"
              # color = "#a32d02",
              ) +
      labs(subtitle = paste0("State: ", input$state)) +
        theme(axis.text.x = element_text(size = 18, angle = 0),
              axis.text.y = element_text(size = 18, angle = 0),
              axis.title = element_text(size = 24, angle = 0)) +
      scale_fill_brewer(palette = "Dark2") +
      scale_x_date(limits = c(as.Date("2020-03-01"), as.Date(Sys.Date())+1),
                   date_breaks = "3 months", date_labels = "%b") +
        guides(fill = F)
    }
      },
    width = 1200, height = 600,
    res = 100
    )

}

shinyApp(ui, server, options = list(height = 1000))
```


# National Overview of Reported Deaths due to Covid-19

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=12, fig.height=6, dpi = 500}


arrow2 <- 
  tibble(
    x1 = c(as_date("2020-05-27")),
    x2 = c(as_date("2020-05-24")),
    y1 = c(1500), 
    y2 = c(1200)
  )


PlotBase +
  geom_bar(data = CovidData %>% filter(deathIncrease >= 0),
           aes(x = date, y = deathIncrease, fill = region),
           stat = "identity", alpha = 1/2, width = 0.7, ) +
  labs(x = "",
       y = "Daily Deaths\ndue to Covid-19\n",
       fill = "Region (per US Census)",
       title = "Covid-19 Deaths by Region\n") +
  annotate("text", x = as_date("2020-06-10"), y = 1600,
           family = "serif", size = 3.5, color = "#a32d02",
           label = "7-Day Running Average \nUS Daily Deaths") +
  geom_curve(data = arrow2, 
             aes(x = x1, y = y1, xend = x2, yend = y2),
             size = 0.3, color = "gray45", curvature = 0.1) +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_line(size = 1/3)) +
  # scale_y_continuous(limits = c(0, 3000))   +
  # coord_cartesian(ylim = c(0, 3000)) +
  geom_path(data = CovidData  %>% group_by(date) %>% summarize(NewDeaths = sum(deathIncrease)),
            aes(x = date,
                y = rollmean(NewDeaths, 7, na.pad=TRUE, align = "right")),
            color = "#a32d02",
            size = 0.75,
            alpha = 2/3) -> DeathStackPlot

DeathStackPlot

# ggsave(filename = "DeathStackPlot.png",
#        plot = DeathStackPlot,
#        device = "png",
#        path = "/home/mhickey/Covid/")

# ggplotly(width = 750, height = 450)
```



New Deaths
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height= 6, dpi = 300}

options(scipen = 999)

Pop <- read_csv("Statepop.csv")

DeathPerCapita <- bak %>%
  filter(!(state %in% c("AS", "MP", "PR"))) %>%
  left_join(Pop,
            by = "state") %>%
  filter(!is.na(Pop)) %>%
  mutate(NewDeathPerCapita = deathIncrease/Pop) %>%
  filter(!is.na(NewDeathPerCapita)) %>%
  filter(NewDeathPerCapita > 0)



Order <- DeathPerCapita %>%
  group_by(state) %>%
  filter(death/Pop == max(death/Pop)) %>%
  ungroup() %>%
  arrange(-death/Pop) %>%
  transmute(state) %>%
  unique()

DeathPerCapita$state <- factor(DeathPerCapita$state, 
                               levels = Order$state)

ggplot() +
  # scale_color_brewer(palette = "Dark2") +
  scale_color_manual(values = c("#a32d02")) +
  facet_wrap(~state, nrow = 3, strip.position="bottom") +
  # scale_y_log10() +
  scale_y_continuous(breaks = c(0, 10, 20, 30, 40)) +
  coord_cartesian(ylim = c(0, 41)) +
  scale_x_date(limits = c(as.Date("2020-02-20"), Sys.Date()+3),
               breaks = seq(lubridate::as_date("2020-03-01"), lubridate::as_date(Sys.Date()+3), 62),
               date_labels = "%b") +
  theme_tufte() +
  labs(y = "Daily New Deaths\nper Million\n",
       x = "",
       fill = "Test Result",
       title = "Death Rates are Reaching new Highs in many US States",
       color = "Trendline (GAM)",
       caption = "Data Source: The Covid Tracking Project\nPlot created by @KeepTheOutliers") +
  theme(axis.text.y = element_text(size = 12, family = "serif"),
        axis.text.x = element_text(size = 8, family = "serif", angle = 90),
        axis.title = element_text(size = 16, family = "serif"),
        legend.title = element_text(size = 10, family = "serif", angle = 0),
        legend.position = "top",
        plot.title = element_text(size = 18, family = "serif", hjust = 0.5),
        plot.subtitle = element_text(size = 12, family = "serif", hjust = 0.5),
        plot.background = element_rect(fill = "#faf8f6"),
        axis.ticks.y = element_blank(),
        panel.grid.minor.y =  element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.margin=unit(c(0.5,0.25,0.25,0.5),"cm"),
        panel.spacing.y = unit(1, "lines"),
        plot.caption = element_text(colour = "grey25", family = "serif", hjust = 0.5, size = 10)) -> DeathPerCapitaPlot

DeathPerCapitaPlot +
  geom_point(data = DeathPerCapita,
            aes(x = date, 
                y = NewDeathPerCapita*1000000),
            size = 1/6, 
            alpha = 1/3,
            color = "grey25") +
  geom_smooth(data = DeathPerCapita,
              aes(x = date,
                  y = NewDeathPerCapita*1000000,
                  color = ""),
              method = "gam",
              alpha = 0,
              # color = "#a32d02",
              size = 1/2,
              na.rm = T
              )  +
  # scale_y_log10(limits = c(0.1, 100)) +
  scale_x_date(limits = c(as.Date("2020-03-01"), as.Date(Sys.Date()))+1,
               date_breaks = "2 months", date_labels = "%b") -> DeathAllPlot


# unique(DeathPerCapita$state) %>% length
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi = 500, fig.width=12, fig.height=6}
#, fig.width=12, fig.height=8, dpi = 300}
# Define UI for app that draws a histogram ----
ui1 <- fluidPage(

  # App title ----
  # titlePanel("Hello Shiny!"),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar panel for inputs ----
     selectInput("state", "State:",
                 c("All US States", as.character(unique(DeathPerCapita$state)) %>% sort()),
                 tableOutput("data")),
    # Main panel for displaying outputs ----
    mainPanel(

      # Output: Histogram ----
      plotOutput(outputId = "distPlot")

    )
  )
)
# Define server logic required to draw a histogram ----
server1 <- function(input, output) {

  # Histogram of the Old Faithful Geyser Data ----
  # with requested number of bins
  # This expression that generates a histogram is wrapped in a call
  # to renderPlot to indicate that:
  #
  # 1. It is "reactive" and therefore should be automatically
  #    re-executed when inputs (input$bins) change
  # 2. Its output type is a plot
  output$distPlot <- renderPlot({

    if(input$state == "All US States") {
      DeathAllPlot
    } else{

      StateDeath <- DeathPerCapita %>%
        filter(state == input$state)

      DeathPerCapitaPlot  +
        geom_point(data = StateDeath,
                   aes(x = date,
                       y = NewDeathPerCapita*1000000),
                   size = 1,
                   alpha = 1/3,
                   color = "grey25") +
        geom_smooth(data = StateDeath,
                    aes(x = date,
                        y = NewDeathPerCapita*1000000,
                        color = ""),
                    method = "gam",
                    alpha = 0,
                    # color = "#a32d02",
                    size = 1.5, alpha = 4/5,
                    na.rm = T
              )  +
        guides(color = F) +
  coord_cartesian(ylim = c(0, 41)) +
        # scale_y_log10(limits = c(0.1, 100)) +
        scale_x_date(limits = c(as.Date("2020-03-01"), as.Date(Sys.Date()))+1,
                     date_breaks = "2 months", date_labels = "%b") +
        theme(axis.text.x = element_text(size = 18, angle = 0),
              axis.text.y = element_text(size = 18, angle = 0),
              axis.title = element_text(size = 24, angle = 0)) +
        labs(subtitle = paste0("State: ", input$state))
    }
      },
    width = 1200, height = 600,
    res = 100
    )

}

shinyApp(ui1, server1, options = list(height = 1000))
```






